---
title: "Getting Started with ipf"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with ipf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>'
)
```

## Overview

Survey samples rarely match the population perfectly on key demographics.
**Raking** (iterative proportional fitting) adjusts case weights so that the weighted sample margins match known population targets.
`ipf` makes this fast and straightforward, with a Rust computational core and a clean R interface.

This vignette walks through a complete raking workflow using the bundled `anes24` dataset — a subset of the 2024 American National Election Study.

## Setup

```{r setup}
library(ipf)
data(anes24)
```

The data contains `r nrow(anes24)` respondents from the ANES 2024 face-to-face sample:

```{r data-overview}
anes24
```

## Step 1: Check levels and missingness

Before writing targets, inspect the levels in your sample and see where values are missing:

```{r levels-and-missing}
table(anes24$sex, useNA = 'ifany')
table(anes24$race, useNA = 'ifany')
table(anes24$income, useNA = 'ifany')
table(anes24$region, useNA = 'ifany')

colSums(is.na(anes24[c('sex', 'race', 'income', 'region')]))
```

When you define targets, the target names must match the data values exactly.
`NA` values are ignored for that variable during raking.

## Step 2: Define population targets

Targets are a named list of named numeric vectors.
Each vector's names must match the levels in your data, and the values should be proportions summing to 1.

```{r targets}
targets <- list(
  sex = c(Male = 0.472, Female = 0.528),
  race = c(
    White = 0.706,
    Black = 0.121,
    Hispanic = 0.107,
    Asian = 0.047,
    Other = 0.019
  ),
  income = c(
    'Under $50k' = 0.151,
    '$50k-$100k' = 0.294,
    'Over $100k' = 0.555
  )
)
```

If your targets don't sum to 1, `ipf` will normalize them automatically with a warning.

## Step 3: Rake

The main function is `rake()`:
```{r rake-basic}
result <- rake(anes24, targets, cap = NULL)
result
```

That's it — `result` is an `ipf_rake` object containing the weights and diagnostics.

If you already have design weights, pass them through `base_weights`:

```{r base-weights}
base_w <- ifelse(anes24$sex == 'Female', 1.1, 0.9)
base_w[is.na(base_w)] <- 1

base_weighted <- rake(anes24, targets, base_weights = base_w, cap = NULL)
base_weighted
```

## Step 4: Inspect results

### Design effect

The design effect measures how much the weighting inflates variance.
A deff of 1.0 means no inflation (uniform weights); higher values mean less effective data.

```{r deff}
design_effect(result$weights)
```

### Per-variable diagnostics

`summary()` shows a full diagnostic report:

```{r summary}
summary(result)
```

The **residual discrepancy** column shows how close the weighted distribution is to the target.

### Tidy output

For programmatic use, the broom-style methods return tibbles:

```{r tidy}
# One row per variable-level
tidy(result)

# One-row summary
glance(result)
```

### Augmenting the data

To use the weights in downstream analyses, attach them to your data:

```{r augment}
weighted_data <- augment(result)
head(weighted_data)
```

The `.weight` column can then be used in downstream analyses.

Here is the regional distribution before and after weighting:

```{r region-comparison}
unweighted_region <- prop.table(table(anes24$region))

weighted_region <- with(
  weighted_data[!is.na(weighted_data$region), ],
  tapply(.weight, region, sum)
)
weighted_region <- weighted_region / sum(weighted_region)

rbind(
  unweighted = round(unweighted_region[names(weighted_region)], 3),
  weighted = round(weighted_region, 3)
)
```

## Advanced options

### Weight bounding

By default, weights are capped at 5.
Tighter bounds reduce extreme weights but can leave more residual mismatch.

```{r cap}
# Unbounded fit from above
range(result$weights)
design_effect(result$weights)

# Default cap
default_bounded <- rake(anes24, targets)
range(default_bounded$weights)
design_effect(default_bounded$weights)

# Tighter cap
tight <- rake(anes24, targets, cap = 3)
range(tight$weights)
design_effect(tight$weights)

# Or specify both min and max bounds
bounded <- rake(anes24, targets, bounds = c(0.3, 3))
range(bounded$weights)
```

### Variable selection

With many potential raking variables, you can let `ipf` select only the most discrepant ones:

```{r var-selection}
targets_many <- list(
  sex = c(Male = 0.472, Female = 0.528),
  race = c(
    White = 0.706,
    Black = 0.121,
    Hispanic = 0.107,
    Asian = 0.047,
    Other = 0.019
  ),
  income = c('Under $50k' = 0.151, '$50k-$100k' = 0.294, 'Over $100k' = 0.555),
  married = c(
    Married = 0.58,
    Widowed = 0.06,
    Divorced = 0.10,
    Separated = 0.01,
    'Never married' = 0.25
  )
)

# Only rake on variables where discrepancy exceeds 5%
result_pct <- rake(anes24, targets_many, type = 'pctlim', pctlim = 0.05)
result_pct$vars_used
```

Use `type = "nlim"` to select the top N most discrepant variables, or `iterate = TRUE` to re-check after each round and add newly discrepant variables.

### Checking discrepancies directly

You can inspect raw discrepancy scores without raking:

```{r discrepancy}
find_discrepant_vars(
  anes24,
  targets_many,
  weights = rep(1, nrow(anes24)),
  choosemethod = 'total'
)
```
